Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-11 vertical-thermal-review.Rmd
Included changes from KAT.doc into .rmd
new version of the manuscript with structuring

### 3.1. Intraspecific trait variation

@@

@@

@@

@@

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003]. 
Early leaf-out allows some tree saplings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019]. Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015]. Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.

### 3.2. Interspecific trait variation

Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes. 
Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022]. 
For instance, across increasing tree size classes, there is an increase in the fractions of species that are deciduous [mirroring the faction of individuals, @conditQuantifyingDeciduousnessTropical2000c; @meakemRoleTreeSize2018]. 
However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.

>>>>

Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in canopy versus understory conditions, thereby shaping community composition across size classes. 
Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022]. 
However, exceptions exist: at least among evergreen species, shade tolerant species tend to have higher LMA than light demanding species, by contrast with the generally higher LMA for sun than shade leaves within canopies [@luskWhyAreEvergreen2008].
Therefore, vertical gradients in $T_{leaf}$ and metabolism within forest ecosystems are underpinned by both intra- and inter-specific variation in traits.

@@

## 3.1 Leaf anatomy and morphology

@@

## 3.2 Metabolic capacity and efficiency

@@

## 3.3 – Light absorption and reflectance

## 3.#. Biochemical protection against foliage light and heat damage

@@

## 3.# Thermal tolerance

"Thermal thresholds for leaves have also been hypothesized to vary with leaf temperature, and thus be higher in sun-exposed upper canopy leaves than in understory leaves. The upper thermal thresholds for leaf survival can be approximated by assessing the photosystem II (PSII) functioning using chlorophyll fluorescence techniques (Krause et al. 2012). Considering sun and shade leaf differences within canopy (Table 1), the temperature at which PSII efficiency declines, Tcrit, and at which it has decreased by 50%, T50, varies modestly across the canopy, being only slightly lower for shade than sun leaves for two of three tree species in Panama (Slot et al., 2019). In Australia, along a vertical gradient within a very open canopy, Acacia papyrocarpa showed greater T50 and higher composite climate stress in the lower, sun-facing canopy than other canopy positions, correlating with low wind speed, greater radiation and Tair, and lower rates of heat dissipation, as indicated by longer leaf thermal time constants (Curtis et al., 2019). These patterns of higher heat tolerance in the warmest leaves are consistent with interspecific patterns across temperature gradients (O’Sullivan et al. 2018; Feeley et al., 2020; Slot et al., 2021a), and experienced maximum temperatures (Perez & Feeley, 2020). However, not enough studies have evaluated thermal tolerances across vertical gradients to make robust inferences [sorry, bit of a blabla sentence to wrap up]."

## 3.# Leaf phenology

The vertical structure of microenvironments likely influences forest photosynthetic capacity [@rey-sanchez_spatial_2016] via leaf phenology. 
In deciduous forests, the timing and extent of seasonal leaf gain and loss is structured across the vertical profile. 
In temperate deciduous forests, seasonal warming occurs earlier in the sheltered understory, facilitating earlier spring bud-break compared to the cooler upper canopy [@augspurgerDifferencesLeafPhenology2003]. 
Early leaf-out allows some tree saplings to do the majority of their annual carbon fixation before canopy foliage and reduces light availability [@leeSpringPhenologicalEscape2021]. 
Within tropical dry-season-deciduous species, the fraction of seasonally deciduous individuals has been found to increase toward larger tree size classes in hotter, brighter environments, with understory individuals typically remaining evergreen [@conditQuantifyingDeciduousnessTropical2000c]. Similarly, on the interspecific level, across increasing tree size classes, there is an increase in the fractions of species that are deciduous (Condit et al., 2000; Meakem et al., 2018). Even in evergreen tropical forests, seasonal variation in leaf quantities follows distinct patterns throughout the vertical profile depending on height and light environments [@smithSeasonalDroughtrelatedChanges2019]. Leaf turnover is faster in the upper canopy than the understory [@osadaLeafDynamicsMaintenance2001], and as a result upper canopy leaves are, on average, younger than shade leaves.
Leaf age distributions affect forest photosynthetic capacity [@wuLeafDevelopmentDemography2016; @niinemetsLeafAgeDependent2016] and stress tolerance via age specific leaf function [@albertAgedependentLeafPhysiology2018; @kikuzawaSynthesisRelationshipsLeaf2006; @mariasImpactsLeafAge2017; @zhouLeafageEffectsTemperature2015]. Leaf phenology thereby both responds to and influences the vertical structure of forest microenvironments and function.
